01-introduction

### 2024-07-03 01-introduction.qmd

In the present paper, we address the role of the perception of meritocracy on the justification of social inequalities by eighth-grade students in Chile, a country characterized by a highly stratified educational system. In particular, we focus in justifying inequalities in health, education and pensions, which are traditional social policy areas where acces to better services could be justified based on payment capacity, referred as to market justice [@lane_market_1986, @lindh_bringing_2023]. Most of the research in this realm has only considered adults, leaving aside the study of how beliefs in this field develop at student age as well as the impact of the school context and the family as the main socialization agencies [@batruch_belief_2022]. Regarding schools, the way in which they deal with unequal conditions of origin has been linked to the *hidden curriculum* [@chafel_schooling_1997], whereby students learn about distributive norms in society and mechanisms of justification of social differences.

@@

The Chilean case is particularly intriguing for studying market justice preferences. This country is characterized by acute and persistent economic inequality, which stands out in Latin America and among OECD countries [@flores_top_2020]. In Chile, the poorest 50% captures only 10% of the total income and has negative wealth, while the richest 1% receives almost 27% of the income and holds 49.6% of the wealth [@chancel_world_2022]. Much of this inequality has been attributed to the deep neoliberal reforms that institutionalized privatization and commodification of various economic sectors. Such reforms where introduced during the dictatorship (1973-1989) and expanded in democracy through concessions, demand credits, and specific regulatory frameworks [@ffrench-davis_reformas_2018]. This shift in economic policy has allowed the unprecedented emergence of markets in areas related to social reproduction, such as health, pensions, and education, with provision and access managed by private entities and segmented by individual payment capacity, heavily reliant on State subsidies [@boccardo_30_2020]. In health, although the majority of the population uses the public insurance system (78.9%), 15.3% are served by private insurers (Observatorio Social, 2023). The pension system is based on individual capitalization, with mandatory contributions managed by private administrators investing in the financial market, currently involving 11 million contributors (Superintendencia de Pensiones, 2024). In education, 30.6% of school enrollment is in public schools, 54.0% in state-subsidized (voucher) private schools, and 9.3% in fully private schools, generally attracting higher-income groups (Ministerio de Educación, 2023).

>>>>

In the present paper, we address the role of the perception of meritocracy on the justification of social inequalities by eighth-grade students in Chile, a country characterized by a highly stratified educational system. In particular, we focus in justifying inequalities in health, education and pensions, which are traditional social policy areas where acces to better services could be justified based on payment capacity, referred as to market justice [@lane_market_1986; @lindh_bringing_2023]. Most of the research in this realm has only considered adults, leaving aside the study of how beliefs in this field develop at student age as well as the impact of the school context and the family as the main socialization agencies [@batruch_belief_2022]. Regarding schools, the way in which they deal with unequal conditions of origin has been linked to the *hidden curriculum* [@chafel_schooling_1997], whereby students learn about distributive norms in society and mechanisms of justification of social differences.

@@

The Chilean case is particularly intriguing for studying market justice preferences. This country is characterized by acute and persistent economic inequality, which stands out in Latin America and among OECD countries [@flores_top_2020]. In Chile, the poorest 50% captures only 10% of the total income and has negative wealth, while the richest 1% receives almost 27% of the income and holds 49.6% of the wealth [@chancel_world_2022]. Much of this inequality has been attributed to the deep neoliberal reforms that institutionalized privatization and commodification of various economic sectors. Such reforms where introduced during the dictatorship (1973-1989) and expanded in democracy through concessions, demand credits, and specific regulatory frameworks [@ffrench-davis_reformas_2018]. This shift in economic policy has allowed the unprecedented emergence of markets in areas related to social reproduction, such as health, pensions, and education, with provision and access managed by private entities and segmented by individual payment capacity, heavily reliant on State subsidies [@boccardo_30_2020]. In health, although the majority of the population uses the public insurance system (78.9%), 15.3% are served by private insurers [@observatoriosocial_estadisticas_2024]. The pension system is based on individual capitalization, with mandatory contributions managed by private administrators investing in the financial market, currently involving 11 million contributors [@superintendenciadepensiones_estadisticas_2024]. In education, 30.6% of school enrollment is in public schools, 54.0% in state-subsidized (voucher) private schools, and 9.3% in fully private schools, generally attracting higher-income groups [@ministeriodeeducacion_resumen_2023].
